Deep Gaussian Processes and Hierarchical Modeling with Functional Data: A Case Study in Cosmological Power Spectra

### 2024-09-21 paper/main.tex
add mean_PCs_oneW plot; update manuscript

The metrics we use to evaluate performance are mean squared error (MSE) and the log score, a proper scoring rule which takes both prediction and intervals into consideration \citep{gneiting2007strictly}. For both metrics, a lower value indicates better performance. Log score and MSE results are shown in Figures \ref{fig:sims_logS} and \ref{fig:sims_MSE} respectively; our DGP.HM model performs favorably against these current benchmark methods.

@@

Here, we begin with $\boldsymbol\eta$, a $n \times m$ matrix composed of $m=111$ posterior means, each of length $n$. To facilitate more efficient estimation, we use singular value decomposition \citep[SVD; e.g.,][]{banerjee2014linear}, where $\boldsymbol\eta = UDV^T$; $U$ and $V$ are orthogonal matrices of size $n \times n$ and  $m\times m$, respectively. $D$ is a diagonal matrix containing the singular values. For this application, $\boldsymbol\eta$ has rank $r=\text{min}\{m,n\}=111$, which determines the number of non-zero elements in $D$. So, we can find an alternative decomposition where $\Sigma$ is a diagonal matrix with the non-zero singular values of $\boldsymbol\eta$, and $U_1$ and $V_1$ are matrices with orthonormal columns of size $n_\eta \times r$ and $m \times r$, respectively.

From here, following modeling strategies from \cite{higdon2008computer, higdon2010estcosmo}, we equivalently decompose $\boldsymbol\eta$ into a principal component (PC) basis matrix $B^* = \frac{1}{\sqrt{m}}U_1\Sigma$ along with its corresponding weights $W^* = \sqrt{m}V_1$. Without loss of generality, we can perform this decomposition after an overall mean trend is removed. If we use $p_\eta < r$ PCs, this will reduce $B$ and $W$ to be the first $p_\eta$ columns of $B^*$ and $W^*$ respectively and result in an approximation where

@@

Figure \ref{fig:mean_PCs_oneW}, illustrates this decomposition with the estimated mean trend, the principal components contained  in $B$, and one example of how the weight for the first PC will vary dependent on the first parameter $\kappa$.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{mean_PCs_oneW.png}
    \caption{\textcolor{red}{UPDATE this from GSMF to spectra!! The estimated mean trend of the GSMF (left), The principal components contained in $B$ (middle), and an illustration of how the weights for the first PC will change as $\kappa$ varies (right)}.}
    \label{fig:mean_PCs_oneW}
\end{figure}

Following \cite{heitmann2009coyote} and \cite{higdon2010estcosmo}, we use GPs to model the principal components' weights; we shall denote this method GP-PC. We assume a powered exponential kernel for the Gaussian process and employ \texttt{GPfit} to perform a multi-start gradient based search for GP's hyperparameters \citep{macdonald2015gpfit}. That is, the $i$th vector of PC weights $w_i(\theta)$, $i \in \{1,\ldots,p_\eta\}$, will be modeled as a zero mean GP: $w_i(\theta) \sim GP(0, \sigma^2R)$, with

@@

We fix $\alpha=1.95$ and estimate each $\beta_k$ with $k \in \{1,...,p_\theta\}$. \textcolor{red}{Replace all $p_\theta, p_\eta, \kappa$; reread and update this section.}

Here we describe the GP-PC model. We discuss how we use the predicted curves that we got previously as training data. \cite{higdon2008computer, higdon2010estcosmo}

\begin{itemize}
    \item show breakdown of estimated mean trend, basis functions/PCs, and weights of a particular PC/BF.
\end{itemize}

@@

Here we show results on held-out cosmologies.  The Cosmic Emu emulator is constructed by (process convolutions, ...). Segue to our method after and compare/contrast with Cosmic Emu.

In addition to predicting for held-out cosmologies, we can also predict over a large uniform or full factorial design to obtain estimated main effects for each parameter in $\theta$, interactions amongst the 8 parameters, and variance decomposition. Should this be included?

@@

Brief recap, directions for future research: BSS-ANOVA in lieu of DGPs, hydrodynamical sims or TCs as other applications of this methodology. A more detailed sensitivity analysis for the parameters, modeling $W$ across different cosmologies, consider modeling for different redshift values other than $z=0$.

@@

The different appendices from dissertation as necessary. Can also put details of dataset in supplementary material. \textcolor{magenta}{Fix references as necessary, e.g. Moran et al. 2024.}

>>>>

The metrics we use to evaluate performance are mean squared error (MSE) and the log score, a proper scoring rule which takes both prediction and intervals into consideration \citep{gneiting2007strictly}. For both metrics, a lower value indicates better performance. Log score and MSE results are shown in Figures \ref{fig:sims_logS} and \ref{fig:sims_MSE} respectively; our DGP.HM model performs favorably against these current benchmark methods. \textcolor{magenta}{If we mention that DGP.HM timings speed up compared to \texttt{deepgp}, we would also need to mention that \texttt{hetGP} is much faster than both of these.}

@@

Here, we begin with a $n \times m$ matrix composed of $m=111$ posterior means, each of length $n$, denoted by $\boldsymbol\eta$. To facilitate more efficient estimation, we use singular value decomposition \citep[SVD; e.g.,][]{banerjee2014linear}, where $\boldsymbol\eta = UDV^T$; $U$ and $V$ are orthogonal matrices of size $n \times n$ and  $m\times m$, respectively. $D$ is a diagonal matrix containing the singular values. For this application, $\boldsymbol\eta$ has rank $r=\text{min}\{m,n\}=111$, which determines the number of non-zero elements in $D$. We find an alternative decomposition where $\Sigma$ is a diagonal matrix with the non-zero singular values of $\boldsymbol\eta$, and $U_1$ and $V_1$ are matrices with orthonormal columns of size $n_\eta \times r$ and $m \times r$, respectively.

Following \cite{higdon2008computer, higdon2010estcosmo}, we equivalently decompose $\boldsymbol\eta$ into a principal component (PC) basis matrix $B^* = \frac{1}{\sqrt{m}}U_1\Sigma$ along with its corresponding weights $W^* = \sqrt{m}V_1$. Without loss of generality, we can perform this decomposition after an overall mean trend is removed. If we use $p_\eta < r$ PCs, this will reduce $B$ and $W$ to be the first $p_\eta$ columns of $B^*$ and $W^*$ respectively and result in an approximation where

@@

Figure \ref{fig:mean_PCs_oneW} illustrates this decomposition with the estimated mean trend, the principal components contained in $B$, and one example of how the weight for the first PC will vary dependent on the first of $p_\theta=8$ cosmological parameters contained in $\theta$.

@@

\begin{figure}
    \centering
    \includegraphics[width=\textwidth]{mean_PCs_oneW.png}
    \caption{The estimated mean trend of the 111 posterior means (left), the principal components contained in $B$ (middle), and an illustration of how the weights for the first PC will change as the first cosmological parameter varies (right) \textcolor{red}{This last plot is wonky because it was made with only the 6 test cosmologies, not a large uniform or FF design, Steve can update this}.}
    \label{fig:mean_PCs_oneW}
\end{figure}

In alignment with \cite{heitmann2009coyote} and \cite{higdon2010estcosmo}, we use GPs to model the principal components' weights; we shall denote this method GP-PC. We assume a powered exponential kernel for the Gaussian process and employ \texttt{GPfit} to perform a multi-start gradient based search for GP's hyperparameters \citep{macdonald2015gpfit}. That is, the $i$th vector of PC weights $w_i(\theta)$, $i \in \{1,\ldots,p_\eta\}$, will be modeled as a zero mean GP: $w_i(\theta) \sim GP(0, \sigma^2R)$, with

@@

We fix $\alpha=1.95$ and estimate each $\beta_k$ with $k \in \{1,...,p_\theta\}$. Using these trained GPs from the PCs, we predict the spectra for unobserved cosmologies.

@@

Here we show prediction results for 6 held-out cosmologies. We compare our method with Cosmic Emu \citep{moran2023mira}, the emulator constructed using the full suite of 111 Mira-Titan model runs. Cosmic Emu uses a Bayesian approach where the smooth spectrum for each cosmology is modeled with a process convolution on Brownian motion. Nonstationarity is permitted through modeling the bandwidth parameter with a process convolution as well; this composition is referred to as a deep process convolution \citep{moran2024dpc}.

Both Cosmic Emu and DGP.HM are capable of modeling a true underlying functions based upon multiple functional realizations, but our approach utilizes DGPs within a hierarchical model in lieu of deep process convolutions. After obtaining the estimated smooth spectra from the training cosmologies, each method utilizes GPs in order to estimate the weights of the PCs from the estimated spectra. Below, we compare prediction performance between Cosmic Emu and DGP.HM for the 6 held-out cosmologies. \textcolor{magenta}{In addition, we can also predict over a large uniform or full factorial design to obtain estimated main effects for each cosmological parameter in $\theta$, interactions amongst the $p_\theta=8$ parameters, and decompose the overall variation by main effects, two-way interactions, and so on. Should this be included?}

Plots for the average root mean squared error (RMSE) for each prediction method over the 6 cosmologies and each $k$ value are shown in Figure \ref{fig:plot_rmse_k}. Comparisons of the spectrum estimation for training cosmologies are shown as well (using RMSE against the weighted average of observations). For predictions, DGP.HM achieves the greatest reduction in RMSE in the region where perturbation theory and the low-resolution runs are deemed unbiased. For the region where only the high-resolution run is unbiased, Cosmic Emu has a lower average RMSE. Over all $k$ values and predictions, DGP.HM has an average RMSE of 0.0089 and Cosmic Emu has an average RMSE of 0.0106. Our RMSE is lower at 56\% of $k$ values considered. The individual predictions for each method and cosmology are shown in Figure \ref{fig:plot_pred_1to6}.

@@

In this work, we propose a novel Bayesian hierarchical model which utilizes DGPs in order to estimate a smooth underlying function from correlated functional observations (as cosmological computer model runs). In addition to the compositional form of the DGP, we include a third component modeling the correlated errors for each spectrum. With all smooth spectra estimated from the training set of 111 batches of model runs, we can predict spectra for unobserved cosmologies.

Multiple avenues for future research emerge: in addition to estimating dark matter power spectra, batches of hydrodynamical simulations exist, and this method could easily be adapted to handle such datasets. Other applications outside of cosmology can be explored, wherever multiple functional outputs exist, perhaps with spatial correlation (for example, observing multiple tropical cyclone forecasts). More detailed sensitivity analyses for the cosmological parameters can be explored to gain insights into how each of these parameters interact, and which are most significant. From a modeling perspective, changes to the model could include modeling the latent layer $W$ jointly across different cosmologies, as well as considering different redshift values other than $z=0$. Leveraging a hybrid model between DGP.HM and Cosmic Emu may also help to obtain accurate estimates on the lower and higher ends of $k$ values, respectively. Additionally, Bayesian smoothing spline ANOVA (BSS-ANOVA) or other methods could be compared in addition to DGPs.

@@

\textcolor{magenta}{The different appendices from dissertation as necessary. Can also put details of dataset in supplementary material.} \textcolor{red}{Fix references as necessary, e.g. Moran et al. 2024.}
